---
title: "CEE218Y Assigment 1"
author: "Awoenam Mauna-Woanya"
date: "1/18/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(censusapi)
library(ggplot2)
library(esri2sf)
library(plotly)

```


In collaboration with Merrick Howarth


The cumulative impacts of historical land use practices on minority populations, directly leading significant wealth gaps today means I am generally in favor of any policy aimed at increasing to opportunity. In this case, this is the opportunity to purchase a home in East Palo Alto. Through debates, whether from the OPA propents at the East Palo Alto City Council Meetings, or anti-OPA websites designed to cajole homeowners against this policy, I've realized that reaching a consensus may not be the mountain I originally thought it was. I believe that the very nature of these discussions can be enhanced with a few key questions: who are the major stakeholders? who would this policy benefit? and why should anyone care?

To understand the impacts of this policy, we must first understand the distinction between the who stakeholders are on paper, and who has a voice. This analysis considers current occupants, whether they are renters or the owners. In the public sphere, there is outward opposition to the OPA through the anti-OPA community website. This website is aimed at convincing homeowners to reject such a policy because it puts their community at-risk. In making this argument, I  assumed that EPA is a majority owner-occupied community, however by using publicly available (albeit through a backdoor method) assesor data, we are able to visualize the following map: 

```{r data, echo=FALSE}
setwd('C:/Users/mouse/OneDrive - Stanford/MS2/WinterQ/218Y/Land Use')
epa_zoning <- read_rds("epa_zoning.rds")


# pulling assessor data or load from save if not first time (should be fine to run from github folder)

smc_exemption <- read_rds("smc_exemption.rds")

# joining parcel data to assessor data
epa_exemption <- smc_exemption %>% 
  mutate(
    # change assessor data's APN to match parcels for joining
    APN = `Parcel Number` %>% 
      str_replace_all("-","")
  ) %>% 
  filter(APN %in% epa_zoning$APN) %>% 
  left_join(epa_zoning) %>% 
  st_as_sf() %>% 
  # filtering to only low density residential housing (single family)
  filter(New_Zone == "R-LD")
```

```{r parcle map, echo=FALSE}

# renter v non renter for year 2019/2020

tenure_map <- leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    data = epa_exemption %>% 
      filter(`Fiscal Year` == "2017/2018")%>% # can change year plotted here
      filter(Exemption %in% c(0)),
    fillColor = "yellow",
    color = "black",
    weight = 0.5,
    group = "Renter Occupied Single Family"
  ) %>% 
  
  addPolygons(
    data = epa_exemption %>% 
      filter(`Fiscal Year` == "2017/2018") %>% # can change year plotted here
      filter(Exemption %in% c(5600,7000)),
    fillColor = "blue",
    color = "black",
    weight = 0.5,
    fillOpacity = 1,
    group = "Owner Occupied Single Family"
  )%>%
  
  addLayersControl(
    overlayGroups = c("Renter Occupied Single Family", "Owner Occupied Single Family"),
    options = layersControlOptions(collapsed = FALSE)
  )

tenure_map
```

Here, we see that this divide on social media and in city council meetings is not present physically in the real-world. In fact, not only are renters and owners neighbors, renters outnumber owner-occupied housing. Realizing that renter and owner occupied housing are mixed together, does not erase the points owners raised, but it sets a precedent that a policy like this is meant to help your neighbor and not a foreign entity; in a way, it could humanize the supporting reasons for OPA. The only catch here could be that non-profits are included in the policy as eligible to buy homes, which could feel more like a foreign entity. 

The next point of consensus is that owners and renters in East Palo Alto are facing similar struggles in housing cost burden, however, the lasting or cumulative impact is worse for renters. By using ACS 5-year census data that measures housing cost as a percentage of income for various income levels for renter and owner occupied households, we can plot the magnitude and prevalence of housing cost burdens. Note that this figure uses 30% as the burden threshold. 

```{r census data, echo=FALSE}
Sys.setenv(CENSUS_KEY = "30f44d95cb0d8c10acfc57d830d39957dbfe8a39")

# acs_vars_2019_5yr <- 
#   listCensusMetadata(
#     name = "2019/acs/acs5",
#     type = "variables"
#   )

# saveRDS(acs_vars_2019_5yr, "acs_vars_2019_5yr.rds")
acs_vars_2019_5yr <- readRDS("acs_vars_2019_5yr.rds")

```

```{r  Renter Housing Burden, echo=FALSE}
# grab EPA place ID; manually
#EPA <- places("06") %>% 
 # filter(NAME == "East Palo Alto")

rent_housingcost <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "place:20956",
    regionin = "state:06",
    vars = "group(B25074)"
  ) %>% 
  mutate(
    cbg = 
      paste0(state, place)
  ) %>% 
  select(!c(GEO_ID, state, place) & !ends_with(c("EA", "MA", "M"))) %>% 
  mutate(
    across(everything(), as.numeric)
  ) %>% 
  pivot_longer(
    ends_with("E"), 
    names_to = "variable", 
    values_to = "estimate"
  ) %>% 
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label, 
    into = c(NA, NA, "house_income", "grossrent_perc"), 
    sep = "!!"
  ) %>% 
  filter(
    !is.na(house_income), 
    !is.na(grossrent_perc)
  )

# we want to filter for people who spend more than 30% of income on rent (regardless of income)

EPA_RentBurden <- 
  rent_housingcost %>% 
  mutate(
    burden_count =
      ifelse(
        grossrent_perc %in% c(
          "30.0 to 34.9 percent",
          "35.0 to 39.9 percent",
          "40.0 to 49.9 percent",
          "50.0 percent or more"
        ),
        estimate,
        NA
      )
  ) %>% 
  group_by(house_income) %>% 
  summarize(
    burdened_sum = sum(burden_count, na.rm = T),
    total_pop = sum(estimate, na.rm = T)
  ) %>% 
  mutate(
    percent_burdened = burdened_sum/total_pop*100
  ) %>% 
  filter(!is.na(percent_burdened)) %>% 
  mutate(
    tenure = "Renter-Occupied Housing"
  )

```


```{r Owner Occupied Burden, echo=FALSE}
owner_housingcost <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "place:20956",
    regionin = "state:06",
    vars = "group(B25095)"
  ) %>% 
  mutate(
    cbg = 
      paste0(state, place)
  ) %>% 
  select(!c(GEO_ID, state, place) & !ends_with(c("EA", "MA", "M"))) %>% 
    mutate(
      across(everything(), as.numeric)
      ) %>% 
  pivot_longer(
    ends_with("E"), 
    names_to = "variable", 
    values_to = "estimate"
  ) %>% 
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label, 
    into = c(NA, NA, "house_income", "gross_housecost_perc"), 
    sep = "!!"
  ) %>% 
  filter(
    !is.na(house_income), 
    !is.na(gross_housecost_perc)
  )

# filter for people who spend more than 30% of their income on housing
EPA_OwnerCostBurden <- 
  owner_housingcost %>% 
  mutate(
    burden_count =
      ifelse(
        gross_housecost_perc %in% c(
          "30.0 to 34.9 percent",
          "35.0 to 39.9 percent",
          "40.0 to 49.9 percent",
          "50.0 percent or more"
        ),
        estimate,
        NA
      )
  ) %>% 
  group_by(house_income) %>% 
  summarize(
    burdened_sum = sum(burden_count, na.rm = T),
    total_pop = sum(estimate, na.rm = T)
  ) %>% 
  mutate(
    percent_burdened = burdened_sum/total_pop*100
  ) %>% 
  filter(!is.na(percent_burdened)) %>% 
  mutate(
    tenure = "Owner-Occupied Housing"
  )

```

```{r Combining Dataframes, echo=FALSE}
EPA_CombinedBurden <- rbind(EPA_OwnerCostBurden, EPA_RentBurden)
  
```


```{r Visualizing Plots,echo=FALSE}

EPA_HousingBurden_Figure <- EPA_CombinedBurden %>% 
  ggplot() +
  geom_bar(
    aes(
      x = house_income %>% factor(levels = c("Less than $10,000:", "$10,000 to $19,999:", "$20,000 to $34,999:", "$35,000 to $49,999:", "$50,000 to $74,999:", "$75,000 to $99,999:", "$100,000 to $149,999:","$100,000 or more:", "$150,000 or more:")),
      y = percent_burdened,
      fill = tenure 
    ),
    stat = "identity",
    position = "dodge"
  )+
  labs(
    x = "Household Income", 
    y = "Percent Housing Burden",
    title = "EPA Percent Housing Cost Burdened by tenure",
    fill = "Tenure"
  ) +
  coord_flip() +
  theme_classic()

EPA_HousingBurden_Figure
```



Segmenting the figure into three sections: Income <\$20,000, \$20,000-\$75,000, >\$75,000, shows two things: first renters and owners both experience housing cost burden to some extent. Within each segment listed above, at least 6% of the population is housing cost burdened. For those making less than $100,000 annually, at least 40% are housing cost burdened. Second, this cost burden is much worse for renters who make around the national average, with an annual income ranging from \$20,000 to \$75,000. For this segment, not only are least 73% are housing cost burdened, but housing expenses are not going towards a property they own, thereby losing wealth-building opportunities. By demonstrating that a policy like OPA could help your neighbors, it might be easier to reach a general consensus.  

Finally, by understanding the recent historical trends, we can show that this policy aligns with recent trends. Also from assessor data, we can show that there is a decreasing number of owner-occupied homes and an increasing number of renter-occupied homes. This could reflect the general rising costs in East Palo Alto acting as a deterrent forcing people to rent instead of buying homes. 

```{r owner_renter_trends, echo= FALSE}

epa_exemption_singleUnit_trend <- epa_exemption %>%
  # adding var to distinguish rented v owned
  mutate(
    Tenure = case_when(
      Exemption %in% c(5600,7000) ~ "Owner Occupied",
      Exemption == FALSE ~ "Rented" #identifying exemptions of $0 as rented
    )
  ) %>%
  
  # removing geometry
  st_set_geometry(
    NULL
  ) %>%
  
  select(
    Tenure, `Fiscal Year`
  ) %>%
  
  group_by(
    Tenure, `Fiscal Year`
  )%>%
  
  # creating count of each tenure for each year
  summarize(`Household Count`=n()) %>%
  
  # adding percentages of each year
  ungroup() %>%
  group_by(`Fiscal Year`) %>%
  mutate(`Percent (%)` = paste0(round(`Household Count`/sum(`Household Count`) * 100,digits=2),"%"))

```

```{r plot_ownership_rent_trend, echo = FALSE}

owner_renter_trend_ggplot <- 
    epa_exemption_singleUnit_trend %>%
  
  # making fiscal year a numeric continuous variable for line chart
  mutate(
    year = case_when(
      `Fiscal Year` == "2015/2016" ~ 2015, 
      `Fiscal Year` == "2016/2017" ~ 2016,
      `Fiscal Year` == "2017/2018" ~ 2017,
      `Fiscal Year` == "2018/2019" ~ 2018,
    )
  ) %>% 
  
  # plot
  group_by(Tenure) %>%
  ggplot() +
  geom_line(
    aes(
      x = year,
      y = `Household Count`,
      color = Tenure,
      group = Tenure,
      text = `Percent (%)`
    )
  ) +
  labs(
    x = "First Identifying Year of Fiscal Year",
    y = "Household Count",
    title = "Owner-Renter Tenure Trends from 2015-2019",
    color = "Tenure Type"
  ) +
  theme_classic()


  # Adding the following renames to proper fiscal year, but removes axis b/c labeling numbers with strings, seems an unavoidable problem b/c either I list x as a factor and the data doesn't plot, I change the labels and the axis disappears, or the axis labels aren't quite right. 

  #scale_x_discrete(breaks = c("2015", "2016", "2017", "2018"),
  #                 labels=c("2015/2016", "2016/2017", "2017/2018", "2018/2019"))


owner_renter_trend_plot <- ggplotly(owner_renter_trend_ggplot, tooltip="text") 


owner_renter_trend_plot

```


The apparent divides on both sides of this debate seems more fabricated as a result of societal expectations from the consequences from a policy like OPA. Anti-OPA folks, stoke fear based on very specific scenarios without realizing that a policy like this that enables more people access to buy homes, could inadvertently support them by improving the selling market. 

The meta-analysis above is built on a more optimistic premise that everyone (renters and homeowners) want everyone to benefit, whereas in reality, most people want what's best for themselves. The example scenarios raised by the anti-OPA groups are valid, however given the requirements needed to qualify for OPA, it would be helpful to have data that shows how many homes would actually qualify. 

Given all of this, I am still in favor of OPA, however I think that we need to do more work to first, give provide the renter community a platform to share the benefits of OPA much like the anti-OPA communty. Second, more directly address the concerns raised by the anti-OPA voices. Even though they are in the minority, they do have a platform and that carries weight. In doing so, both populations can hopefully move towards the consensus that we are creating opportunity for everyone, both their neighbors and themselves. 


