---
title: "amw_highway_notes"
author: "Awoenam Mauna-Woanya"
date: "5/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(readxl)
library(tigris)
library(sf)
library(leaflet)
library(jsonlite)
library(esri2sf)
library(mapview)
library(mapboxapi)

Sys.setenv(CENSUS_KEY="30f44d95cb0d8c10acfc57d830d39957dbfe8a39")
pa_api <- "6C39C4C7-9A53-11EC-B9BF-42010A800003" 
path <- 'G:/Shared drives/S22 218Z/Air Quality (OCOB)/Data/'
participants <- read_csv(paste0(path,'participants.csv'))

```

# Highway Geometry

Grab highway shape geometry from https://opendata.mtc.ca.gov/datasets/san-francisco-bay-region-roadways/explore?location=37.490972%2C-122.217834%2C13.66 

```{r, include=FALSE}
bay_highways_shp <- readRDS(paste0(path, "bay_highways_shp.rds"))
```

```{r}
# map it
mapview(bay_highways_shp)
```



# Grab Indoor/Outdoor Sensors

Grab all our sensors, both in our study and across the Bay (filtered to San Mateo County). Highways are merged via union for upcoming analysis.

```{r, include=FALSE}
# grab all bay sensors
bay_sensors <- readRDS(paste0(path,"Sensors/bay_sensors_2022-04-17.rds"))

# grab our 27 sensors
all_27_sensors <- readRDS(paste0(path, "Sensors/all_27_sensors.rds"))

# SMC boundary
smc_boundary <- counties("CA", cb = T) %>% 
  filter(NAME == "San Mateo") %>% 
  st_transform(4326)

# Pair all bay highways to SMC
smc_highways <- bay_highways_shp %>% 
  st_transform(4326) %>% 
  .[smc_boundary, ] %>% 
  st_union()

# SMC only sensors instead of all bay sensors
smc_sensors <- bay_sensors %>% 
  st_transform(4326) %>% 
  .[smc_boundary,]


```


```{r}

mapview(all_27_sensors) + mapview(smc_highways)

```


# Indoor/Outdoor Analysis within 1/2 mile of highways

Add a .5 mile buffer to highways to qualitatively compare air quality for sensors near and far from highways. .5 miles chosen arbitrarily.

```{r, include=FALSE}
# create a 1/2 mile buffer of the SMC highways
highway_buffer <- smc_highways %>% 
  st_transform(26910) %>% 
  st_buffer(800) %>% 
  st_transform(4326)
  
smc_sensors_near_highway <- smc_sensors %>% 
  .[highway_buffer,]


```

```{r}
mapview(smc_sensors_near_highway) + mapview(smc_highways)
```


# add tags for upcoming visualizations (near highway/sensor type)

```{r, include=FALSE}
all_27_near_highway <- 
  all_27_sensors %>% 
  .[highway_buffer,]

all_27_sensors <- 
  all_27_sensors %>% 
  mutate(
    near_highway = case_when(
      sensor_index %in% all_27_near_highway$sensor_index ~ "YES",
      TRUE ~ "NO"
    ),
    type = "participant"
  )

smc_sensors <- 
  smc_sensors %>% 
  mutate(
    near_highway = case_when(
      sensor_index %in% smc_sensors_near_highway$sensor_index ~ "YES",
      TRUE ~ "NO"
    ),
    type = "original"
  ) %>% 
  rbind(
    all_27_sensors
  )
```

# Visualize sensors near highway map

```{r}
leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolylines(
    data = smc_highways,
    fillColor = "black", 
    fillOpacity = 1,
    color = "black", 
    weight = 2
  ) %>% 
  addPolygons(
    data = highway_buffer,
    fillColor = "yellow",
    fillOpacity = .2,
    color = "yellow",
    weight = .2
  ) %>% 
  addCircleMarkers(
    data = smc_sensors %>% filter(near_highway == "YES", type == "original"),
    fillColor = "green",
    fillOpacity = 1,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .8,
    group = "Original Sensor Near from Highway"
  ) %>% 
  addCircleMarkers(
    data = smc_sensors %>% filter(near_highway == "NO", type == "original"),
    fillColor = "red",
    fillOpacity = .5,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .2, 
    group = "Original Sensor Away from Highway"
  ) %>% 
  addCircleMarkers(
    data = all_27_sensors %>% filter(near_highway == "YES", type == "participant"),
    fillColor = "purple",
    fillOpacity = 1,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .8,
    group = "Participant Sensor Near Highway"
  ) %>%
  addCircleMarkers(
    data = all_27_sensors %>% filter(near_highway == "NO", type == "participant"),
    fillColor = "purple",
    fillOpacity = 1,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .2,
    group = "Participant Sensor Away Highway"
  ) %>% 
  addLayersControl(
    overlayGroups = c("Original Sensor Away from Highway",
                   "Original Sensor Near from Highway",
                   "Participant Sensor Near Highway",
                   "Participant Sensor Away Highway"),
    options = layersControlOptions(collapsed = FALSE)
  )


# mapshot(m, file = paste0(path, "highway_analysis.png"))

```


# Binary AQI analysis

```{r, include=FALSE}
# call Purple Air API
json <- fromJSON(paste0(
    "https://api.purpleair.com/v1/sensors?api_key=",
    pa_api,
    "&fields=name,location_type,latitude,longitude,pm2.5_1week,temperature,humidity,primary_id_a,primary_key_a,secondary_id_a,secondary_key_a,primary_id_b,primary_key_b,secondary_id_b,secondary_key_b"
  ))

# pull all recent data
all_sensors <- json %>% 
  .$data %>% 
  as.data.frame() %>% 
  set_names(json$fields) %>% 
  filter(
    !is.na(longitude),
    !is.na(latitude)
  ) %>% 
  st_as_sf(coords = c("longitude","latitude"), crs = 4326) %>% 
  mutate(location_type = ifelse(
    location_type == 0,
    "outside",
    "inside"
  ))

# subset to SMC
smc_pa_sensors <- all_sensors %>% 
  .[smc_boundary,]

smc_sensors_clean <- smc_pa_sensors %>% 
  filter(
    !is.na(pm2.5_1week),
    !is.na(humidity)
  ) %>% 
  mutate(
    PM25 = 0.524*as.numeric(pm2.5_1week) - 0.0852*as.numeric(humidity) + 5.72,
    AQI = case_when(
      PM25 <= 12 ~ 
        paste(round(50/12*PM25), "Good"),
      PM25 <= 35.4 ~ 
        paste(round((100-51)/(35.4-12)*(PM25 - 12) + 51), "Moderate"),
      PM25 <= 55.4 ~
        paste(round((150-101)/(55.4-35.4)*(PM25 - 35.4) + 101), "Moderately Unhealthy"),
      PM25 <= 150.4 ~
        paste(round((200-151)/(150.4-55.4)*(PM25 - 55.4) + 151), "Unhealthy"),
      PM25 <= 250.4 ~
        paste(round((300-201)/(250.4-150.4)*(PM25 - 150.4) + 201), "Very Unhealthy"),
      TRUE ~ 
        paste(round((500-301)/(500.4-250.5)*(PM25 - 250.5) + 301), "Hazardous")
    )
  ) %>% 
  separate(
    AQI,
    into = c("AQI","AQI_Cat"),
    sep = " ",
    extra = "merge"
  ) %>% 
  mutate(
    AQI = as.numeric(AQI),
    AQI_Cat = AQI_Cat %>% factor(levels = c("Good", "Moderate","Moderately Unhealthy","Unhealthy","Very Unhealthy","Hazardous"))
  )

# add near highway tag
smc_sensors_complete  <- 
  smc_sensors_clean %>% 
    mutate(
      near_highway = case_when(
        sensor_index %in% smc_sensors_near_highway$sensor_index ~ "YES",
        TRUE ~ "NO"
      )
    )
# saveRDS(smc_sensors_complete, paste0(path,"smc_sensors_AQI.rds"))


```

```{r, include=FALSE}
smc_sensors_AQI <- readRDS(paste0(path,"smc_sensors_AQI.rds"))
```

```{r, include=FALSE}
SMC_summary <- smc_sensors_AQI %>% 
  group_by(near_highway) %>% 
  summarize(
    average_AQI = mean(AQI)
  )


```

The plot below shows little difference between AQi in sensors near the highway or further away.
```{r}
AQI_near_highway_plot <- 
  ggplot(data = SMC_summary,
         aes(x = near_highway,
             y = average_AQI)) +
  geom_col() + 
  scale_fill_brewer(palette = "YlOrRd")

AQI_near_highway_plot
```

# Improve analysis to measure AQI continuously
Use st_nearest to determine the distance from each sensor to the nearest highway. Below is a map that shows one sensor connected to its nearest highway.

```{r}

nearest_distance <- 
  st_nearest_points(smc_sensors_AQI[1,],smc_highways) %>% 
  st_sf()

# visualize
leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolylines(
    data = smc_highways,
    fillColor = "black", 
    fillOpacity = 1,
    color = "black", 
    weight = 2
  ) %>% 
  addPolygons(
    data = highway_buffer,
    fillColor = "yellow",
    fillOpacity = .2,
    color = "yellow",
    weight = .2
  ) %>% 
  addCircleMarkers(
    data = smc_sensors_AQI %>% filter(near_highway == "YES"),
    fillColor = "green",
    fillOpacity = 1,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .8,
    group = "Original Sensor Near from Highway"
  ) %>% 
  addCircleMarkers(
    data = smc_sensors_AQI %>% filter(near_highway == "NO"),
    fillColor = "red",
    fillOpacity = .5,
    color = "black",
    opacity = 1,
    radius = 6,
    weight = .2, 
    group = "Original Sensor Away from Highway"
  ) %>% 
  addPolylines(
    data = nearest_distance %>% 
      st_transform(4326),
    color = "blue",
    weight = .5
  ) %>% 
  addLayersControl(
    overlayGroups = c("Original Sensor Away from Highway",
                   "Original Sensor Near from Highway"),
    options = layersControlOptions(collapsed = FALSE)
  )


# distance_to_hwy <- 
#   st_nearest_points(smc_sensors_AQI[3,ncol(smc_sensors_AQI)],smc_highways) %>% 
#   st_length() %>% 
#   as.numeric()


```

```{r, include=FALSE}
# use map_dbl to get the distance from each point to the nearest highway list
smc_distance_to_hwy <- 
  1:nrow(smc_sensors_AQI) %>% 
  map_dbl(function(x){
    
    smc_sensors_AQI[x, ncol(smc_sensors_AQI)] <- 
      st_nearest_points(smc_sensors_AQI[x, ],smc_highways) %>% 
      st_length()
    
  })

smc_distance_to_hwy <- smc_distance_to_hwy %>% 
  floor()

smc_sensors_AQI$`d_to_hwy (m)` <- smc_distance_to_hwy



```

# Regression analysis to see relationship between AQI and distance

First, plotting AQI relative to distance from highway, then a linear regression summary showing that according to our model, distance from highway is a poor predictor of AQI.
```{r}
hwy_AQI_distance <- ggplot(smc_sensors_AQI,
                           aes(`d_to_hwy (m)`,AQI)
                           ) +
  geom_point() +
  geom_smooth()

hwy_AQI_distance

# linear regression formula
fit <- lm(AQI ~ `d_to_hwy (m)`, data = smc_sensors_AQI)
summary(fit)

```

For every meter increase in distance to highway, our model predicts a -4.175e-5 decrease in AQI. 


